Renewal and stability in populations structured by remaining years of life

### 2014-09-11 ThanoRepro/ThanoRepro.tex
intro and eq punctuation (not done)

\documentclass{article}
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{tikz}
%\usepackage[active,tightpage]{preview}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
\usepackage{nth}
% for the d in integrals
\newcommand{\dd}{\; \mathrm{d}}

@@

\onehalfspacing
All demographic forces vary over age in known and regular ways. Information on
such forces, along with the size and age structure of a population allows the
demographer to make predictions about the future size and structure of a population. In this
paper, I explore some formal demographic consequences of a particular
redefinition of age. Instead of counting age as the time passed between birth and the present (or
some other moment), consider age as the amount of time left from the present
until death\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure.}. Individuals in this case move in the same
direction along imaginary life lines, but the reference point for age is now at
the end of the life line instead of at the beginning. For individuals, the
timing of events in life is mirrored exactly between the two definitions of age,
but aggregate patterns differ.

@@

Temporal perspectives and rescalings of the lifecourse are abundant in
demography already. It is my view that these efforts have enriched the
discipline by producing new insights and questions much more than they have obscured it by
producing conundrums. Likely some demographic phenomena are best
described as a function of time since birth, and others of time until death,
while still others can be a function of both to some degree. Fertility and
reproduction are plausibly a function of both age perspectives, and the question
of how to partition such forces and events over time is a new conundrum that I
do not tackle here. The model presented here should therefore not be interpreted
as a suggestion for a better way to model or project population, but as an
alternative way to tell the story of population renewal.

Thanatological (counting down towards death) and chronological (counting up
from birth) renewal models, such as the Lotka-Euler renewal model, both
approximate the movement of population stocks over time, but from opposite
vantage points. Specifying a model of thanatological renewal expands the toolbox
available to approach complex temporal population structure. I present a set of
preliminary findings to help understand the implications of temporal structure.

@@

A lifetable approach to transforming chronologically classified data into
thanatologically classified data is valid when the given lifetable accurately
describes the future mortality of age-classified population. To make
a projective statement about a current stock of population, one does well to
factor in future changes in mortality, and to allow the trajectory of changes
to vary between ages/cohorts. In the case of a stable population model,
mortality is assumed invariant over time, which means that the same survival
pattern underlies each birth cohort. This is the case we explore, and so the
following formulas assume a single lifetable.

The basic transformation to thanatological age proceeds by defining a
deaths density function, similar the lifetable $d(a)$, but over two
dimensions: chronological age attained, $a$, and remaining years of life, $y$.
This function we denote $d(y|a)$ --- The density distribution of thanatological
age depends on chronological age. I use the same function name $d()$, because
it is essentially the same notion as the familiar $d(a)$, but rescaled as a
function of chronological age. For ages higher than 0, simply condition
on survival:\footnote{In the reliability literature, our concept of $d(y|a)$ is denoted by \begin{equation} X - t | X > t
\end{equation}
where $X$ is the total lifespan and $t$ is the current chronological age
attained.}

\begin{equation}
\label{eq:vaupel1}
d(y | a) = \mu(a+y)\frac{l(a+y)}{l(a)}
\end{equation}

@@

\begin{equation}
\label{eq:dyadiscret}
d_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \quad\quad \text{,}
\end{equation}

\noindent where $\omega$ is the highest
age.\footnote{Equation~\eqref{eq:dyadiscret} ill incur some error depending on
the method used to derive $L_a$. To remove this error, rescale over the $a$ margin so that each distribution conditioned on $a$
sums to 1. Working with $l_a$ rather than $L_a$ eliminates error and yields
similar results, but is not necessarily better.} One produces in this way a
probability density function over future death times for each chronological age,
$a$. The $d(y|a)$- weighted average of $y$ for a given $a$ returns the familiar
remaining life expectancy column of the lifetable, $e(a)$. If $P(a)$ is a
chronological age-classified population count, we can derive $P(y)$, a
thanatological age-classified population count using ~\eqref{eq:transform}.

\begin{align}
\label{eq:transform}
P(y) =& \int_0^\infty P(a) d(y | a) \dd a
\end{align}

@@

It bears repeating that lifetable-derived thanatological age structure is probabilistic until after death, since it refers to the future according to an assumed mortality pattern.
As such, Figure~\ref{fig:USrecomp} is projective in nature, a forward-looking
glance at population attrition given the population stock and lifetable of a
particular moment, whereas Figure~\ref{fig:USdecomp} is reflective in nature,
since a population's chronological age-structure is mostly the fruit of past
fertility, but also migration and to a lesser degree attrition. Any count
classified by chronological age can be reclassified in this way given a
corresponding lifetable. Exploiting this, we derive thanatological fertility
rates, $f^\star(y)$, by applying \eqref{eq:transform} to age-classified birth
counts, $B(a)$, to get $B(y)$ and again to exposure-to-risk, $E(a)$ (here we
take exposures from all ages), to get $E(y)$ and then dividing:

\begin{equation}
f^\star(y) = \frac{B(y}{E(y)}
\end{equation}

@@

\FloatBarrier
\section*{The thanatological renewal model}
For the renewal model that follows, we are most interested in using the kind
of fertility rates shown in Figure~\ref{fig:FySpaghetti}, and we refer to a
closed, unisex population based on female vital rates. Assuming constant vital rates, the
births for the present year are given by 
\begin{align}
B(t) =& \int _0^\infty P(y,t)f^\star(y) \dd y = \int _0^\infty P(a,t)f(a) \dd a
\quad\text{,} \intertext{where $a$ indexes chronological age, $y$ indexes
thanatological age, $P(a), P(y)$ are population counts, and $f^\star(y), f(a)$ are
exact specific fertility probabilities (rates), which here are best thought of
as the fertility rates of females born to females. The thanatological integral can
be broken down back in terms of chronological age:} =& \int_{y=0}^\infty
\int_{a=0}^\infty f^\star(y)\frac{P(a,t)d(a+y)}{l(a)} \dd a \dd y \quad\text{,}
\intertext{where $d(a)$ is the continuous lifetable death distribution with
radix of 1, so $d(a)/l(0)$. We can relate the present population to past births
with $P(a,t) = B(t-a)l(a)$:}
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t-a)d(a+y)\dd a \dd y
\quad\text{.} \intertext{Eventually strong ergodicity will assert itself, and $B(t)$ will be related to
$B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth:}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\quad\text{.}
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation:}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) d(a+y)e^{-ra}\dd a \dd y
\quad\text{.}
\end{align}
Now compare this to Lotka's chronological formulation
\begin{align}
\label{eq:lotka}
1 =& \int_{a=0}^\infty f(a)l(a)e^{-ra}\dd a \quad\text{,}
\intertext{which can be expressed as well as:}
  =& \int_{a=0}^\infty \int_{y=0}^\infty f(a)d(a+y)e^{-ra}\dd y \dd a
  \label{eq:lotka2}
\end{align}
and note that \eqref{eq:lotka2} and \eqref{eq:thanoren} are really quite
similar, since $\int _{y=0}^\infty d(a+y)\dd a = l(a)$. For intuition, notice that $l(a)$ is here split
up into pieces of $d(a)$, and imagine a 2D surface of these, where one axis is chronological age and the
other axis is thanatological age. For the chronological case \eqref{eq:lotka},
we multiply chronological age-specific fertility rates, $f(a)$, over the
chronological age margin, and for the thanatological case we multiply
$f^\star(y)$ over the thanatological age margin.

@@

Once one finds $r$ from \eqref{eq:thanoren}, other familiar stable population
parameters can be calculated. For instance, we may calculate the mean
thanatological generation time, $T^\star$:
\begin{equation}
\label{eq:Ty}
 T^\star =  \frac{\int _{y=0}^\infty \int _{a=0}^\infty y e^{-ra} d(a+y)
 f^\star(y) \dd a \dd y}{\int _{y=0}^\infty \int _{a=0}^\infty e^{-ra} d(a+y)
 f^\star(y) \dd a \dd y} \quad \text{.}
\end{equation}
Literally, this is the mean of the remaining lifespans of new mothers in the
stable population. The net reproduction rate, $R_0$ is related by, e.g.,
\begin{equation}
\label{eq:R0fromTy}
R_0 = e^{r T^\star}
\end{equation}
The birth rate, $b$, is given by
\begin{align}
\label{eq:eybrate}
b =& \frac{1}{\int _{y=0}^\infty \int _{a=0}^\infty e^{-ra} d(a+y) \dd a \dd
y}\quad\text{,}
\intertext{the denominator of which is the inside of
equation~\eqref{eq:thanoren} less $f^\star(y)$. This simplifies to} =&
\frac{1}{ \int _{a=0}^\infty e^{-ra} l(a) \dd a } \quad \text{,}
\intertext{which is exactly the same as that valid for
chronological age in the Lotka-Euler setup, except that $r$ here comes from the
thanatological renewal model. If the stable age structure, $c^\star(y)$, is
known, then we may take the sum of hypothetical births over thanatological age:}
=& \int _{y=0}^\infty f^\star(y) c^\star(y) \dd y \quad{.}
\end{align}

The stable thanatological age structure, $c^\star(y)$, is the proportion of the stable population with remaining years to live $y$:
\begin{align}
\label{eq:cy}
c^\star(y) =& b \int _{a=0}^\infty e^{-ra} d(a+y) \dd a 
\intertext{or, given the stable chronological age structure, $c(a)$, and a
lifetable,} =& \int _{a=0}^\infty c(a)d(y|a) \dd a \label{eq:cyfromca}\\ \notag
\\
           =& \frac{\int _{a=0}^\infty d(a+y)e^{-ra}\dd a}{\int
           _{a=0}^\infty l(a)e^{-ra}\dd a}  \quad\text{.}
\end{align}

>>>>

\documentclass{article}
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{tikz}
%\usepackage[active,tightpage]{preview}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
\usepackage{nth}
% for the d in integrals
\newcommand{\dd}{\; \mathrm{d}}
\newcommand{\ec}{\quad\quad\text{,}}
\newcommand{\ep}{\quad\quad\text{.}}

@@

\onehalfspacing
All demographic forces vary over age in known and regular ways. Information on
such forces, along with the size and age structure of a population, allows the
demographer to make predictions about the future size and structure of a population. In this
paper, I explore some formal demographic consequences of a particular
redefinition of age. Instead of counting age as the time passed between birth and the present (or
some other moment), consider age as the amount of time left from the present
until death\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure.}. The first kind of age is called
chronological age and the kind of age explored in this paper is
called thanatological age. Individuals in this case move in the same direction
along imaginary life lines, but the reference point for age is now at the end of the life line instead of at the beginning. For individuals, the timing of events in life is mirrored exactly between the two definitions of age,
but aggregate patterns differ.

@@

Temporal perspectives and rescalings of the lifecourse are abundant in
demography already. It is my view that these efforts have enriched the
discipline by producing new insights and questions much more than they have obscured it by
producing conundrums. Likely some demographic phenomena are best
described as a function of time since birth, and others of time until death,
while still others can be a function of both to some degree. Fertility and
reproduction are plausibly a function of both age perspectives, and the question
of how to best partition such forces and events over time is a new conundrum
that I do not tackle here. Instead I describe some of the formal consequences
of the Lotka-Euler renewal model when viewed from the thanatological perspective
of age. The renewal model presented here is not necessarily a suggestion for a
better way to model or project population, but in the minimal sense it
represents an alternative way to tell the story of population renewal.

The chronological-age classified Lotka-Euler renewal model and the
thanatological renewal model both approximate the movement of population stocks
over time, but from opposite vantage points. Specifying a model of thanatological renewal expands the toolbox
available to approach complex temporal population structure. The next section
of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph. I present a set of preliminary empirical findings to help
understand the implications of temporal structure.

@@

A lifetable approach to transforming chronologically classified data into
thanatologically classified data is valid when the given lifetable accurately
describes the future mortality of age-classified population. To make
a projective statement about a current stock of population, one does well to
factor in future changes in mortality, and to allow the trajectory of changes
to vary over ages and cohorts. In the case of a stable population model,
mortality is assumed invariant over time, which means that the same survival
pattern underlies each birth cohort. This is the case we explore, and so the
following formulas assume a single lifetable pattern.

The basic transformation to thanatological age proceeds by defining a
deaths density function, similar the lifetable $d(a)$, but over two
dimensions: chronological age attained, $a$, and remaining years of life, $y$.
This function we denote $d(y|a)$ --- The density distribution of thanatological
age conditional on chronological age. We use the same function name $d()$,
because it is essentially the same notion as the familiar $d(a)$, but rescaled as a
function of chronological age. For ages higher than 0, simply condition
on survival:\footnote{In the reliability literature, our concept of $d(y|a)$ is denoted by \begin{equation} X - t | X > t
\end{equation}
where $X$ is the total lifespan and $t$ is the current chronological age
attained.}

\begin{equation}
\label{eq:vaupel1}
d(y | a) = \mu(a+y)\frac{l(a+y)}{l(a)} \ep
\end{equation}

@@

\begin{equation}
\label{eq:dyadiscret}
d_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \ec
\end{equation}

\noindent where $\omega$ is the highest
age.\footnote{Equation~\eqref{eq:dyadiscret} ill incur some error depending on
the method used to derive $L_a$. To remove this error, rescale over the $a$ margin so that each distribution conditioned on $a$
sums to 1. Working with $l_a$ rather than $L_a$ eliminates error and yields
similar results, but is not necessarily better.} One produces in this way a
probability density function over future death times for each chronological age,
$a$. The $d(y|a)$- weighted average of $y$ for a given $a$ returns the familiar
remaining life expectancy column of the lifetable, $e(a)$. If $P(a)$ is a
chronological age-classified population count, we can derive $P(y)$, a
thanatological age-classified population count using

\begin{align}
\label{eq:transform}
P(y) =& \int_0^\infty P(a) d(y | a) \dd a \ep
\end{align}

@@

It bears repeating that lifetable-derived thanatological age structure is probabilistic until after death, since it refers to the future according to an assumed mortality pattern.
As such, Figure~\ref{fig:USrecomp} is projective in nature, a forward-looking
glance at population attrition given the population stock and lifetable of a
particular moment, whereas Figure~\ref{fig:USdecomp} is reflective in nature,
since a population's chronological age-structure is mostly the fruit of past
fertility, but also migration and to a lesser degree attrition. Any count
classified by chronological age can be reclassified in this way given a
corresponding lifetable. Exploiting this, we derive thanatological fertility
rates, $f^\star(y)$, by applying \eqref{eq:transform} to age-classified birth
counts, $B(a)$, to get $B(y)$ and again to exposure-to-risk, $E(a)$ (here we
take exposures from all ages), to get $E(y)$ and then dividing

\begin{equation}
f^\star(y) = \frac{B(y}{E(y)} \ep
\end{equation}

@@

\FloatBarrier
\section*{The thanatological renewal model}
For the renewal model that follows, we are most interested in using the kind
of fertility rates shown in Figure~\ref{fig:FySpaghetti}, and we refer to a
closed, unisex population based on female vital rates. Assuming constant vital rates, the
births for the present year are given by 
\begin{align}
B(t) =& \int _0^\infty P(y,t)f^\star(y) \dd y = \int _0^\infty P(a,t)f(a) \dd a
\ec \intertext{where $a$ indexes chronological age, $y$ indexes
thanatological age, $P(a), P(y)$ are population counts, and $f^\star(y), f(a)$ are
exact specific fertility probabilities (rates), which here are best thought of
as the fertility rates of females born to females. The thanatological integral can
be broken down back in terms of chronological age:} =& \int_{y=0}^\infty
\int_{a=0}^\infty f^\star(y)\frac{P(a,t)d(a+y)}{l(a)} \dd a \dd y \ec
\intertext{where $d(a)$ is the continuous lifetable death distribution with
radix of 1, so $d(a)/l(0)$. We can relate the present population to past births
with $P(a,t) = B(t-a)l(a)$:}
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t-a)d(a+y)\dd a \dd y \ep 
\intertext{Eventually strong ergodicity will assert itself, and $B(t)$ will
be related to $B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth:}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\ep
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation:}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) d(a+y)e^{-ra}\dd a \dd y
\ep
\end{align}
Now compare this to Lotka's chronological formulation
\begin{align}
\label{eq:lotka}
1 =& \int_{a=0}^\infty f(a)l(a)e^{-ra}\dd a \ec
\intertext{which can be expressed as well as}
  =& \int_{a=0}^\infty \int_{y=0}^\infty f(a)d(a+y)e^{-ra}\dd y \dd a \ec
  \label{eq:lotka2}
\end{align}
and note that \eqref{eq:lotka2} and \eqref{eq:thanoren} are really quite
similar, since $\int _{y=0}^\infty d(a+y)\dd a = l(a)$. For intuition, notice that $l(a)$ is here split
up into pieces of $d(a)$, and imagine a 2D surface of these, where one axis is chronological age and the
other axis is thanatological age. For the chronological case \eqref{eq:lotka},
we multiply chronological age-specific fertility rates, $f(a)$, over the
chronological age margin, and for the thanatological case we multiply
$f^\star(y)$ over the thanatological age margin.

@@

Once one finds $r$ from \eqref{eq:thanoren}, other familiar stable population
parameters can be calculated. For instance, we may calculate the mean
thanatological generation time, $T^\star$:
\begin{equation}
\label{eq:Ty}
 T^\star =  \frac{\int _{y=0}^\infty \int _{a=0}^\infty y e^{-ra} d(a+y)
 f^\star(y) \dd a \dd y}{\int _{y=0}^\infty \int _{a=0}^\infty e^{-ra} d(a+y)
 f^\star(y) \dd a \dd y} \ep
\end{equation}
Literally, this is the mean of the remaining lifespans of new mothers in the
stable population. The net reproduction rate, $R_0$ is related by, e.g.,
\begin{equation}
\label{eq:R0fromTy}
R_0 = e^{r T^\star}
\end{equation}
The birth rate, $b$, is given by
\begin{align}
\label{eq:eybrate}
b =& \frac{1}{\int _{y=0}^\infty \int _{a=0}^\infty e^{-ra} d(a+y) \dd a \dd
y}\ec
\intertext{the denominator of which is the inside of
equation~\eqref{eq:thanoren} less $f^\star(y)$. This simplifies to} =&
\frac{1}{ \int _{a=0}^\infty e^{-ra} l(a) \dd a } \ec
\intertext{which is exactly the same as that valid for
chronological age in the Lotka-Euler setup, except that $r$ here comes from the
thanatological renewal model. If the stable age structure, $c^\star(y)$, is
known, then we may take the sum of hypothetical births over thanatological age:}
=& \int _{y=0}^\infty f^\star(y) c^\star(y) \dd y \ep
\end{align}

The stable thanatological age structure, $c^\star(y)$, is the proportion of the stable population with remaining years to live $y$
\begin{align}
\label{eq:cy}
c^\star(y) =& b \int _{a=0}^\infty e^{-ra} d(a+y) \dd a \ec
\intertext{or, given the stable chronological age structure, $c(a)$, and a
lifetable,} =& \int _{a=0}^\infty c(a)d(y|a) \dd a \label{eq:cyfromca}\\ \notag
\\
           =& \frac{\int _{a=0}^\infty d(a+y)e^{-ra}\dd a}{\int
           _{a=0}^\infty l(a)e^{-ra}\dd a}  \ep
\end{align}
